```{r}
library(ggplot2)
library(tidyverse)
loan_data=read.csv('prosperLoanData.csv')
loan_data <- loan_data[,c(2,5,6,8,16,17,18,19,20,21,22,26,27,29,30,32,33,34,35,36,37,38,41,42,43,47:62,64,67,68,77)]
loan_data <- mutate(loan_data, credit_score=(CreditScoreRangeLower+CreditScoreRangeUpper)/2)
```


First I look at the distribution of loan amounts:

```{r}
ggplot(loan_data,mapping = aes(x = LoanOriginalAmount))+
+     geom_histogram(bins=10)
```

It looks like most loans are between \$2,500 and \$5,000. The distribution is skewed to the right, with a few outlier loans valued at $30,000 and above.

I increase the number of bins now to see a more detailed distribution: 

```{r}
ggplot(loan_data,mapping = aes(x = LoanOriginalAmount))+
+     geom_histogram(bins=50)
```

It looks like the most common loan amount is somewhere around \$4,000, and that many loans come in round numbers or just about (hence the spikes at the \$10,000,\$15,000,\$20,000, and \$25,000 marks).

Next I look at how the amount loaned varies with homeownership status. First let's see how many homeowners versus renters borrow money:

```{r}
ggplot(loan_data)+
+     geom_bar(aes(IsBorrowerHomeowner),width=.4)
```

There isn't much difference between the counts, so I don't think I need to convert to proportions for the next chart:

```{r}
ggplot(loan_data)+
+     geom_histogram(mapping = aes(x=LoanOriginalAmount,fill=IsBorrowerHomeowner),bins=50)
```

Just for the sake of it, this is how the data would look like if we converted to proportions (note the density variable on the y-axis).

```{r}
ggplot(loan_data,aes(x=LoanOriginalAmount,stat(density),fill=IsBorrowerHomeowner))+geom_density(alpha=.3)
```

Next, I display average loans based on income:

```{r}
loan_data %>% group_by(IncomeRange) %>% summarize(avg_loan=mean(LoanOriginalAmount)) %>% arrange(desc(avg_loan)) %>% ggplot()+geom_bar(aes(x=reorder(IncomeRange,avg_loan),y=avg_loan),stat='identity')
```

Not surprisingly, loan amounts are higher the more income a borrower has. The same point can be seen if we look at the distribution of loan amounts by income:

```{r}
ggplot(loan_data)+
+     geom_freqpoly(aes(x=LoanOriginalAmount,color=IncomeRange),size=1,bins=10)
```
Looking at the income levels, it seems strange that there is an income range called '$0', especially when 'Not employed' (and 'Not displayed' is also available). Let's look closer at datapoints with this level:

```{r}
loan_data %>% filter(IncomeRange=='$0') %>% head(20)
```

It looks like every borrower with this level has an unverified income, and further that dent-to-income data is not available. Let's make sure this is right. It turns out there are only 23 entries where income is \$0 and income can be verified, and there are no entries where income is \$0 and debt-to-income is populated. 
```{r}
loan_data %>% filter(IncomeRange=='$0'& IncomeVerifiable!='False') %>% count()
```

```{r}
loan_data %>% filter(IncomeRange=='$0'& DebtToIncomeRatio!='NA') %>% count()
```
However, there are whose income is not verifiable even if income is a valid value:
```{r}
loan_data %>% filter(IncomeRange!='$0'& IncomeVerifiable=='False') %>% count()
```

Furthermore, debt-to-income data is available for every borrower with a non-\$0 income:
```{r}
loan_data %>% filter(IncomeRange!='$0'& DebtToIncomeRatio=='NA') %>% count()
```

This is just something to keep in mind if I do anything with debt-to-income information.

Going back to the relation between income categories and loan count or amount, unemployed persons and people whose income is listed as \$0 made far fewer loans than other income categories:

```{r}
loan_data$IncomeRange <- factor(loan_data$IncomeRange,levels=c("Not displayed","Not employed","$0","$1-24,999","$25,000-49,999","$50,000-74,999","$75,000-99,999","$100,000+"))
ggplot(loan_data)+
+     geom_bar(aes(x=IncomeRange))
```

This is no surprise: given what we have seen before about \$0 income category, it sounds like these are just borrowers whose income could not be verified (and where there is no data on their debt-to-income ratio). It makes sense there will be few people like this whose loan applications will be accepted. In fact, since it's not clear what income these borrowers really have, I will leave them aside for now (I won't take them out of the data frame however, since these data points could be useful for other questions).

However, I'll look in more detail at the other category that is less likely to get loans - the unemployed:
```{r}
loan_data %>% filter(IncomeRange=='$0') %>% ggplot()+geom_histogram(aes(x=LoanOriginalAmount))
```

```{r}
loan_data %>% filter(IncomeRange=='Not employed'&LoanOriginalAmount>5000) %>% group_by(ListingCategory..numeric.) %>% summarise(count=n()) %>% arrange(desc(count))
```

Most of the higher loans incurred by unemployed persons were for the purpose of debt consolidation and business (codes 1 and 3, respectively) - although one loan is to buy/fix a boat, and another is for an engagement ring. As an aside, this strikes me as a bad classification, since unemployed people might have any level of income from sources other than a job (which might explain the boat).

##Heading

I turn next to the analizing the relationship between income range and debt-to-income ratio.
```{r}
ggplot(loan_data)+geom_freqpoly(aes(x=DebtToIncomeRatio,color=IncomeRange),bins=100)
```
This is not very illuminating because the outliers make the main part of the chart unreadable. We can see however a spike at value 10 (meaning 'has a debt 10 times or higher than income'!), which seems to be elicited by people in the lowest income range. Also the warning above the chart is entirely expected - these are the borrowers who don't have a valid DTI entry. (That's also why the \$0 category is missing completely from the chart). We can also try to get more insight into the distributions of DTI based on income range by using boxplots:
```{r}
ggplot(loan_data)+geom_boxplot(aes(x=IncomeRange,y=DebtToIncomeRatio))
```

This graph is not very useful in terms of differences in the median (any differences here are dwarfed by outliers) but it does suggest a huge difference regarding the highest quartiles in different income levels: even high DTI higher income borrowers are very close to the median - see for example the folks making over \$100,000, who are mostly clustered below DTI=1.25. In contrast, around 25% of borrowers who make under \$25,000 seem to have a DTI higher than that!

Let's 'zoom in' where most of the data points are, and remember to look at the outliers later on. Let's also change the y-axis from count to density, since in this case there is a significant difference in the number of people borrowing money in each income range.
```{r}
ggplot(loan_data)+
  geom_freqpoly(aes(x=DebtToIncomeRatio,color=IncomeRange,stat(density)),size=1,bins=100)+
  coord_cartesian(xlim=c(-.5,1.25))
```

People in lower income ranges have higher debt-to-income ratios, at least when outliers are excluded (the spikes in the 'not employed' frequency is probably due to the lower number of people in this catgegory).There are some people who have a negative debt-to-income ratio - not sure what that means. 

Now let's look at the most extreme outliers:
```{r}
ggplot(loan_data)+geom_freqpoly(aes(x=DebtToIncomeRatio,color=IncomeRange),bins=100)+coord_cartesian(xlim=c(9.75,10.25),ylim=c(0,500))
```
As expected, most people with the highest DTI are in the lowest income category.

A big caveat here is that we cannot really derive conclusions about the relationship between income level and DTI for the population as a whole. Since DTI is an important factor in deciding whether an applicant will get a loan, our dataset is probably skewed towards people who have better-than-average DTI ratios (One plausible possibility though is that the DTI gap between lower and higher income people is even higher for the population as a whole - since a higher income person with a high DTI is more likely to get a loan - and so appear in our data - than a lower income person with the same high DTI).

##Heading

Next I look at loan category. First I add a column to the tibble that translates loan category code:

```{r}
loan_codes <- tibble(code=0:20,y=c( 'Not Available','Debt Consolidation', 'Home Improvement', 'Business', 'Personal Loan', 'Student Us', 'Auto', 'Other', 'Baby&Adoption', 'Boat', 'Cosmetic Procedure', 'Engagement Ring', 'Green Loans','Household Expenses', 'Large Purchases', 'Medical/Dental', 'Motorcycle', 'RV', 'Taxes', 'Vacation','Wedding Loans'))

loan_data <- loan_data %>% left_join(loan_codes,by=c('ListingCategory..numeric.'='code')) %>% rename('loan_desc'='y')
```

Let's look at the highest average amounts of loan by loan type:
```{r}
loan_data %>% group_by(loan_desc) %>% summarise(mean=mean(LoanOriginalAmount)) %>% ggplot()+geom_bar(aes(x=reorder(loan_desc,mean),y=mean),stat='identity')+coord_flip()
```

And let's look at the most common kinds of loan:

```{r}
loan_data %>% group_by(loan_desc) %>% summarise(count=n()) %>% ggplot()+geom_bar(aes(x=reorder(loan_desc,count),y=count),stat='identity')+coord_flip()
```

The vast majority of loans are intended for debt consolidation. And it looks like this is the dominant category in each income range:
```{r}
loan_data %>% group_by(loan_desc,IncomeRange) %>% summarise(count=n()) %>% ggplot()+geom_bar(aes(x=reorder(loan_desc,count),y=count,fill=IncomeRange),stat='identity')+coord_flip()
```

#Credit Rating versus Loan Size (for completed/defaulted loans)

For loans that have been completed or defaulted, I plot credit rating against loan size:

```{r}
loan_data %>% filter(LoanStatus %in% c('Completed','Defaulted')) %>% ggplot()+geom_point(aes(x=jitter(credit_score),y=LoanOriginalAmount,color=LoanStatus),alpha=1/3)
```

#Homeownership versus default rate 

I look next at homeownership versus default rate. Because I'm interested in the default _rate_, the chart will need to visualize proportions rather than raw counts.
```{r}
loan_data %>% filter(LoanStatus %in% c('Completed','Defaulted')) %>% ggplot()+geom_bar(aes(x=IsBorrowerHomeowner,fill=LoanStatus),position='fill',width=.4)
```

The default rates are almost the same, however, homeowners defaulted at a slightly lower rate than renters.

#Homeownership versus late pay

Similarly, we can look at the effect homeownership has on late payments. I create a new column that indicates if a borrower is late with a payment or not (this applies only to current loans). Then I display the percentage of late payments for homeowners versus non-homeowners. ('1' indicates borrower is late, '2' indicates payment is on time.)

```{r}
loan_data <- loan_data %>% mutate (late=ifelse(LoanStatus %in% c('Past Due (>120 days)','Past Due (1-15 days)','Past Due (16-30 days)','Past Due (31-60 days)','Past Due (61-90 days)','Past Due (91-120 days)'),1,0),not_late=ifelse(LoanStatus %in% c('Current','FinalPaymentInProgress'),2,0),borrower_late=late+not_late)

loan_data$borrower_late <- as.character(loan_data$borrower_late)

loan_data %>% filter(borrower_late >0) %>% ggplot()+geom_bar(aes(x=IsBorrowerHomeowner,fill=borrower_late),position='fill')
```


#Homeownership versus credit score and DTI

```{r}
ggplot(loan_data)+geom_point(aes(x=jitter(credit_score),y=DebtToIncomeRatio,color=IsBorrowerHomeowner),alpha=.2)
```

#Employment duration versus income range

```{r}
ggplot(loan_data)+geom_point(aes(x=EmploymentStatusDuration,y=StatedMonthlyIncome*12),alpha=.2)+
+     coord_cartesian(xlim=c(0,360),ylim=c(0,200000))+geom_hline(yintercept=mean(loan_data$StatedMonthlyIncome*12),color='red',size=2)
```
It looks like people who stayed longer in the same job tend to make closer to the average wage. 
